Algebraic Groups\vspace{-1ex}

### 2020-02-28 Algebraic Geometry/Algebraic-Geometry.tex
First update in a while

>>>>

\section{February 21, 2020}
Today we are going to talk about sheaves of modules. Let $(X,\O_X)$ be a ringed space. Then 
\begin{defn}
	An $\O_X$-module (``sheaf of $\O_X$ modules'') is a sheaf of abelian groups $\calF$ along with a morphism 
	\[\O_X\to \iHom(\calF,\calF)\]
	of sheaves of rings. Equivalently it is an action 
	\[\O_X\times\calF\to \calF\]
	or (locally) $\calF(U)$ is an $\O_X(U)$-module and the ``experience is functorial.''

	A homomorphism of $\O_X$-modules is a morphism $\varphi:\calF\to\calG$ of abelian sheaves that is compatible with the $\O_X$-action.
\end{defn}
\begin{lem}
	The category $\lmod{\O_X}$ of $\O_X$ modules is an abelian category.
\end{lem}

\section{February 24, 2020}
We talked about the sheaf of sections of a vector bundle on a manifold. I didn't write it down but it is a good thought.
Here are some

\subsection{Moar sheavz}
Let $X=\Spec A$ and let $M\in \lmod A$. We create a sheaf of modules $\tilde M$ by saying $\tilde M(D(f))=M_f$. 
The right thing to do is $B=\prod A_{f_i}$. Then max did a bunch of shit. The idea is that this is almost exactly the 
same construction that you use to create the structure sheaf. There is some cochain stuff going on here that is quite cool.

So the functor $\tilde -:\lmod A\to \lmod{\O_X}$ is a functor and is exact, commutes with $\otimes,$ and satisfies $\Gamma(\Spec A,\tilde M)=M$ and 
for all $p\in X$, $\tilde M_p=M_p$. It is also fully faithful but \textbf{not essentially surjective}.

To show this last idea, consider the space with two points: $R= k[[t]]$ and let $U\subset\Spec R$ be the open point. THen define 
a $\O_{K(R)}$ module on the localization $K(R)$ at $U$. Just let this sheaf be $\tilde K(R)$. But then consider the 
\[\Gamma(\Spec R,\tilde K(R))=0\quad\text{and}\quad\Gamma(K(R),\tilde K(R))=K(R)\]
which doesn't arise as the ``sheafification'' (DON'T read too much into this word although it may kinda be true) of an $R$-module.

\subsection{(Quasi-)Coherent sheaves}
Hey, don't read Hartshorne for coherent sheaves, apparently.

\begin{defn}
	Let $X$ be a scheme. Then an $\O_X$ modules $\scrF$ is \textbf{quasicoherent} if there exists an open affine cover $U_i=\Spec A_i$ and modules $M_i\in\lmod{A_i}$
	such that for all $i$, $\scrF|_{U_i}=\tilde M_i$.
\end{defn}
Better:
\begin{defn}
	$\scrF$ is quasicoherent if and only if there exists an open (not necessarily affine!) cover $U_i\subseteq X$ such that there exists a presentation 
	\[\O_{U_i}^{\oplus I}\to \O_{U_i}^{\oplus J}\to \scrF|_{U_i}\to 0\]
	for all $i$ and for some index sets $I$ and $J$.
\end{defn}

The right thing to read is FAC by Serre where he does this whole thing (in French).

\subsection{What does coherence mean?}
Quasicoherence gives us $\calO^I\to\calO^J\to \scrF\to 0$ is exact. We can say that $\scrF$ is quasicoherent finitely genreated if $J$ can be taken 
to be finite. It is quasicoherent finitely presented if both $I$ and $J$ can be taken to be finite.

A sheaf is \textbf{coherent} if for ANY finite $J$ and surjection $\O^J\to \scrF$, $\ker(\pi)$ is finitely generated.
In the Noetherian case, all of these coincide.

WARNING! Coherence is not stable under pullback. So what kind of perations can we do? IF we have $f:X\to Y$ is a map of ringed spaces, we get the pullback $f_\ast:\O_X$-mod$\to\O_Y$-mod and we want to have a left 
adjoint $f^\ast$. Recal that we have $f^\sharp:\O_Y\to f_\ast\O_X$, where the codomain acts on $f_\ast\scrF$.

How do we define the adjoint? $f^\sharp$. Wow I just stopped in the middle of a thought.

\section{February 26th, 2020}
We had a functorial experience. We showed that these adjoint maps $f_\ast,f^\ast$ exist on the level of (abelian)
sheaves on $X$ and $Y$ and also they exist (THOUGH ARE DIFFERENT) when we're talking about $\O_X$ and $\O_Y$ modules. 
These also restrict to maps on the quasicoherent sheaves (and finitely presented ones) but not, in general, to the coherent sheaves.

\begin{thm}
	The map $\tilde -:\lmod{A}\to \mathbf{QCoh}_{\Spec A}$ is an equivalence of categories.
\end{thm}
\begin{cor}
	If $X$ is quasi-compact and quasi-separated, then for all $f\in \Gamma(X,\O)$ and $s\in \scrF(D(f))$, and $\scrF\in \mathbf{QCoh}_X$,
	there exists a positive integer $n$ such that $f^ns$ extends to a global section $\sigma\in\Gamma(X,\scrF)$. In other words, 
	all poles of $s$ have finite order.
\end{cor}

\subsection{Next time...}
\begin{prop}
	Let $f:X\to Y$ be a morphism of schemes. Then 
	\begin{enumerate}
		\item $f^\ast\mathbf{QCoh}_Y\subseteq\mathbf{QCoh}_X$ (we already showed this).
		\item If $f$ is quasi-compact, then $f_\ast\mathbf{QCoh}_X\subseteq\mathbf{QCoh}_Y$.
	\end{enumerate}
\end{prop}

### 2020-02-28 Algebraic Groups/Algebraic-Groups.tex
First update in a while

>>>>

\section{February 21st, 2020}
We did a lot of work last time to show that $k[G]$ was injective and furthermore that every module injects into 
an injective module. In general, finding projectives is much harder.

\subsection{Projectives}
\begin{lem}
	Let $E$ be a simple $G$-module. Then there exists an injective hull $Q_E$ such that 
	\[E\hookrightarrow Q_E\]
	with $\operatorname{soc}Q_E=E$ and $Q_E$ is indecomposable.
\end{lem}
\begin{rmk}
	Any injective in $\Rep G$ is a direct sum of indecomposables.
\end{rmk}
\begin{lem}
	Let $P$ be a projective $G$-module. Then $P$ is injective.
\end{lem}
\begin{prf}
	Let $P$ be a projective module. We claim that for any $G$-module $M$, $P\otimes_k M$ is again projective. This follows from tensor-hom adjointness.
	So $\Hom(-,P)$ is exact. Let 
	\[0\to M_1\to M\to M_2\to 0\]
	be a short exact sequence of finite dimensional modules. Dualizing is exact, as is $-\otimes P$, so 
	\[0\leftarrow M_1^\ast\otimes P\leftarrow M^\ast\otimes P\leftarrow M_2\otimes P\leftarrow 0\]
	is exact. Thus $M\otimes P$ splits as $(M_1\otimes P)\oplus M'$.

	But then notice that 
	\[\Hom_G(M_1,P)\simeq(\Hom_k(M,P))^G\simeq(M_1^\ast\otimes P)^G\]
	but then $(M^\ast\otimes P)^G$ splits as 
	\[(M_1^\ast\otimes P)^G\oplus M'^G\]
	and this proves surjectivity.
\end{prf}
\begin{rmk}
	Notice that this alst proof could possibly be better. The finite dimensionality part could be able to be weakened/dropped by
	avoiding the tensor product representation.
\end{rmk}

\subsection{Frobenius categories}
The upshot to all of this is that 
\begin{prop}
	If there exists a projective module $P$ in $\Rep G$, then the collections of injectives and projectives are the same.
\end{prop}
\begin{prf}
	Let $P$ be projective. Let $P_1\subseteq P$ be a finite-dimensional submodule. Then 
	\[M\hookrightarrow M\otimes \End_k(P_1)\simeq M\otimes P_1^\ast\otimes P_1\hookrightarrow M\otimes P_1^\ast\otimes P\]
	which we have seen is both injective and projective.

	So take a simple $M$ with injective hull $Q_M$. Then we get maps $Q_M\to M\otimes P_1^\ast\otimes P$ and $M\otimes P_1^\ast\otimes P\to Q_M$ 
	by leveraging injectivity and projectivity. Thus $Q_M$ is a direct summand of $M\otimes P_1^\ast\otimes P$, which is projective. 
	A little clean up gets us the result.
\end{prf}

\begin{rmk}
	When $P$ is injective iff projective, the category $\Rep G$ is a Frobenius category.
\end{rmk}

\begin{rmk}
	FOr smooth algebraic groups $\Gm$, $\Ga$, and $\GL_n$, there are no projective modules.
	If $G$ is finite (finite groups, Frob kernels ${\Ga}_{(r)}$, $\mu_p$, ${\GL_n}_{(r)}$) we have that the collections of projective and injectives are the same.
\end{rmk}
\begin{rmk}
Domkin put together a complete list of affine group schemes with projective modules.
\end{rmk}

\subsection{Cohomology}
We know now that $\Rep G$ has enough injectives, so if $\calF:\Rep G\to \Vectk$ is left exact we can take an 
injective resolution $M\to I^\bullet$ and then 
\[(\calR^n\calF)(M)\eqdef H^n(\calF(I^\bullet)).\]

Recall the Grothendieck spectral sequence:  Let $\calF:\calC\to \calD$ and $\calF:\calD\to \calE$. Then say 
we want to compute the derived functors of $\calF'\circ\calF$. Assume that $\calF'$ is left exact and that 
$\calF$ takes injectives to $\calF'$ acyclic objects. Then we have a spectral sequence whose second page is 
\[E_2^{m,n}=\calR^m\calF'\circ\calR^n\calF\Rightarrow \calF^{m+n}(\calF'\circ\calF)\]

\begin{rmk}
	Notice that in the Grothendieck spectral sequence if $\calF'$ is exact (so $\calR^n\calF=0$ when $n>0$),
	you get 
	\[\calF'\circ\calR^n\calF\xrightarrow{\sim}\calR^n(\calF'\circ\calF)\]

	You can do the same thing with exactness of $\calF$.
\end{rmk}

\subsubsection{Notation}
We will use the following notation:
\begin{align*}
	\Ext_G^n(M,-)=\calR^n(\Hom_G(M,-))\\
	H^n(G,M)=\Ext_G^n(k,M)\\
	H^n(G,k)=\Ext^n_G(k,k)\\
	H^\ast(G,k)=\oplus_0^\infty\Ext_G^n(k,k)
\end{align*}
and this last object is called the cohomology ring of $G$.

\subsubsection{Applications}
\begin{prop}
	We have the derived adjointness 
	\[\Ext_G^n(M\otimes V,N)\simeq \Ext_G^n(M,\Hom(V,N))\]
\end{prop}

\section{February 24th, 2020}
Today we are going to talk more about the Grothendieck spectral sequence.
\subsection{Examples}
Let's begin by proving 35.5.2
\begin{prf}
	Let $\calF(-)=M\otimes -$ and $\calF'(-)=\Hom_G(-,N)$. Then 
	\[\calF'\circ\calF(V)=\Hom(M\otimes V,N)\]
	and so $\scrR^n(\calF')\circ \calF(V)=\Ext^n(M\otimes V,N)$
	but we know that this is isomorphic to $\scrR^n(\calF'\circ\calF)(V)$,

	I need to work this out once to understand the details. It involved using the regular tensor-hom identity and 
	the convergence of the Grothendieck spectral sequence again.
\end{prf}

\begin{prop}
	\[\Ext_G^n(M,\scrR^m\Ind_H^G N)\Rightarrow \Ext_H^{n+m}(m,N)\]
\end{prop}
\begin{prf}
	Leverage Frobenius reciprocity and use the fact $\scrF=\Ind_H^G$ is left exact and takes injectives to injectives. Then if we let $\scrF'=\Hom_G(M,-)$,
	we get $\scrF'\circ\scrF(N)=\Hom_H(M,N)$ and it pops out.
\end{prf}

\begin{defn}
	We say a subgroup scheme $H\le G$ is \textbf{exact in $G$} if $\Ind_H^G$ is exact. Equivalently if 
	$\scrR^{n>0}\Ind_H^G=0$.
\end{defn}
\begin{cor}
	If $H$ is exact, we have an isomorphism 
	\[\Ext_G^n(M,\Ind_H^G N)\xrightarrow{\sim}\Ext_H^n(M,N)\]
\end{cor}
This is, of course, a much easier thing to work with. But an interesting question is when $H$ is exact.
\begin{prop}
	Let $k$ be a field. Then $H$ is exact if and only if $k[G]$ is an injective $H$ module.
\end{prop}
\begin{rmk}
	The direction ``$\Leftarrow$'' works when $k$ is a commutative ring.
\end{rmk}
\begin{rmk}
	$k[G]=\Ind_1^G k=\Ind_H^G(\Ind_1^H k)=\Ind_H^G(k[H])$
\end{rmk}
\begin{prf}
	Take a short exact sequence $0\to M_1\to M\to M_2\to 0$ of $H$-modules. Then we do the old argument:
	take a splitting of $M\otimes k[G]$ and we get that $M_1\otimes k[G]$ is a summand of $M\otimes k[G]$.
	Then applying invariants preserves this decomposition.

	The rest of this (and all of these properties) live in Jantzen chapter 1 section 4.
\end{prf}

\begin{rmk}
	If $H$ is a finite groups scheme, then $H$ is exact (of everything)! The motivation for this: say we have $H\le G$ 
	any two affine group schemes.

	In analogy, there is a statement about Hopf algebras (see my notes for that). If $A\hookrightarrow B$ is a map of two Hopf algebras (perhaps if $B$ is cosemisimple?)
	This idea ports over here somehow. It can be proved via sheaf cohomology.
\end{rmk}

Next we're going to try to write down the ``generalized version'' of the tensor identity using the Grothendieck spectral sequence.
\[(\scrR^n\Ind_H^G)\circ(-\otimes \Res^G_H N)\cong (\scrR^n\Ind_H^G(-))\otimes N\]

\section{February 26th, 2020}
This is all from Jantzen chapter I section 5. Apparently it is a bit hand-wavy.
\subsection{Associated sheaves}
Our goal is to describe $\scrR^n\Ind_H^G(M)$ as $H^n(G/H,\calL(M))$ where $\calL(M)$ is a sheaf on $G/H$.
We are going to think of $\scrL:\Rep H\to \Sh G/H$ as a functor.

\subsubsection{Quotients}
The history of these quotients go back to Demazure and Gabriel. The idea is that we don't genreally have 
quotients in $\Sch$, so we extend to the \textbf{sheaves in the fppf topology}. This is a Grothendieck topology that stands for ``
finitely presented faithfully flat''. It's French, don't question it.

When we have an algebraic group $G$ acting on a scheme $X$, we want to know when the quotient $X/G$ in this broader context 
is actually a scheme.

Let us assume that $G$ acts freely (pointwise). We assume that $k$ is a field. We can drop this, but then we have 
to carry around the assumption of flatness everywhere. Let $X$ be a scheme of finitel type (algebraic) over $k$ and $G$ an 
affine group scheme (also of finite type). Then we have 
\begin{prop}
	If $G$ acts freely and $X/G$ is a scheme, then 
	\[X(R)/G(R)\hookrightarrow (X/G)(R)\]
\end{prop}
\begin{prop}
	($G$ always acting freely) If $X/G$ is a scheme, there exists an isomorphism 
	\[X\times G\xrightarrow{\sim}X\times_{X/G}X\quad\text{via}\quad (x,g)\mapsto (x,xg)\]
	where on the level of points, this is $\{(x,x')|x=x'g\}$ (we are using right actions obviously).
\end{prop}
From here on out, we are going to assume that $X/G$ is a scheme unless we say something to the contrary.
\begin{ex}
	$\P n(R)$ is, as a set, the set of free rank one direct summands of $R^{n+1}$. Work out how (and if!) this is true.
	Then we can define 
	\[\A{n+1}_\ast(R)=\{(a_0,\dots,a_n)\in R^{n+1}|\sum Ra_i=R\}\]
	which is a subscheme of $\A{n+1}$. Now we claim that 
	\[\A{n+1}_\ast/\Gm\xrightarrow{\sim}\P{n}\]
	And we had some trouble making this precise. This is in Jantzen, however.
\end{ex}
You can do similar things with the Grassmann schemes $\operatorname{Gr}(k,n)$.
\begin{thm}
	Let $G$ be a finite group scheme and $X$ an affine scheme with a free $G$ action. Then $X/G$ is an affine scheme with 
	coordinate ring $k[X]^G$.
\end{thm}
\begin{thm}
	Let $H\le G$ be algebraic group schemes (sometimes requires that $H$ is closed, but this is automatic when we look at it from the functorial 
	perspective and require that $H$ is an algebraic subgroup scheme). Then (with $H$ acting on the right of $G$) $G/H$ (or in the other case $H\setminus G$) is a scheme.
	In particular, $G\times H\simeq G\times_{G/H}G$.
\end{thm}

\section{February 28th, 2020}
\subsection{Associated Sheaves}
The set up is the following: we have a scheme $X/k$, algebraic group scheme $G/k$ acting freely on $X$ such that $X/G$ a scheme.
We are constructing a functor 
\[\calL:\Rep G\to \Sh X/G\]
and actually to $\O_{X/G}$ modules. Let $\pi:X\to X/G$ be affine, flat and algebraic.
\begin{thm}
	\begin{enumerate}
		\item If $Y\subset X/G$ is affine (open), then $\pi^{-1}Y\subset X$ is affine (open). T
		\item If $Y\subset Y/G$ is affine open, then we get the map $\pi:\pi^{-1}Y\to Y$ induces a $k[Y]$ structure on $k[\pi^{-1}Y]$ such that it is a faithfully flat, finitely presented $k[Y]$ module.
	\end{enumerate}
\end{thm}

Now we construct things in the following way. Let $M$ be a $G$-modules. Then for any open $U\subset Y$, define 
\[\calL(M)(U)=\left(M\otimes_k k[\pi^{-1}U]\right)^G\]
where we put the diagonal action on the tensor product. Notice $\pi^{-1}(U)\subset X$.

Why is this the right thing to do? Notice that 
\[M\otimes k[\pi^{-1}U]=\operatorname{Mor}(\pi^{-1}U,M_a)=\{f:\pi^{-1}U\to M_a|f(xg)=g^{-1}f(x)\}.\]
We claim that $\calL(M)$ is a sheaf, and specifically a sheaf of $\O_{X/G}$-modules. That is, the map $\pi:X\to X/G$ gives us 
a $\O_{X/G}$ module structure on $\O_X$ via $\pi^{ast}:\pi^{-1}\O_{X/G}\to \O_X$. On any $U$ we have, for 
$f_1\in \O_{X/G}(U)$ and $f\in\calL(M)(U)$, an element 
\[f_1\cdot f\in\calL(M)(U).\]

\begin{prop}
	\begin{enumerate}
		\item $\calL$ is a sheaf of $\O_{X/G}$ modules.
		\item $\calL$ is exact.
		\item $\calL(M)$ is quasi-coherent.
		\item If $M$ is a finite dimensional $G$-module, $\calL(M)$ is coherent.
	\end{enumerate}
\end{prop}
We may talk about a proof of this next time.
\begin{ex}
	Let $M=k$. Then 
	\[\calL(k)(U)=k[\pi^{-1}U]^G=\{f:\pi^{-1}U\to k_a=\A1|f(xg)=f(x)\}\]
	so we are just looking for invariant functions. The claim is that this is actually 
	\[\calL(k)(U)=\{f:\pi^{-1}U/G\to \A1\}=\{f:U\to A1\}=k[U]=\O_{X/G}(U)\]
\end{ex}
We want to know what happens to injective modules. All of our injectives are going to look like $M\otimes k[G]$ or direct summands of this or will be 
otherwise comparable to this. So if we have $\pi:X\to X/G$,
\[\calL_{X/G}(M\otimes k[G])\simeq \pi_\ast\calL_{X/e}(M)\]
This is requires some thought and computation.

\begin{prop}
	\begin{enumerate}
		\item The functor sending 
		\[M\mapsto H^0(X/G,\calL(M))=\Gamma(\calL(M))\]
		is left exact (but not right exact!)
		\item If $X$ is affine, then the derived functors are 
		\[M\mapsto H^n(X/G,\calL(M))\]
	\end{enumerate}
\end{prop}
\begin{prf}
	The second part looks almost obvious from the Grothendieck spectral sequence. We need, however, 
	that $\calL$ takes injective $G$-modules to acyclic sheaves. Why is that true? Take $M\otimes k[G]$ an 
	injective $G$-module, then we saw that 
	\[\calL(M\otimes k[G])\simeq \pi_\ast\calJ_{X/e}(M).\]
	We need to check that 
	\[H^n(X/G,\pi_\ast\calL_{X/e}(M))\]
	vanishes for all $n\ge 1$. THere is a problem in Hartshorne that gives this to us for an affine morphism between separated (something else) schemes.
\end{prf}
\begin{prop}
	Let $H\le G$. Then 
	\begin{enumerate}
		\item $\calR^n\Ind_H^G(M)\simeq H^n(G/H,\calL_{G/H}(M))$.
		\item If $G/H$ is Noetherian, then $\scrR^n\Ind_H^G(M)=0$ for $n>\dim G/H$.
		\item If $G/H$ is projective, then $\scrR^n\Ind_H^G(M)$ is finitely generated whenever $M$ is finite dimensional.
		\item If $G/H$ is affine, $\Ind_H^G$ is exact.
	\end{enumerate}
\end{prop}
\begin{prf}
	For the first statement, We are just taking higher derived functors of 
	\[H^0(GH,\calL(M))=\calL(M)(G/H)=(M\otimes k[G])^H=\Ind_H^G(M)\]
	where the isomorphisms here are natural. So the derived functors are the same!
\end{prf}
\begin{cor}
	If $H$ is a finite group scheme, then $\Ind_H^G$ is exact.
\end{cor}
\begin{prf}
	When know that $G/H$ is affine and in fact is $\Spec k[G]^H$.
\end{prf}
